---
title: "Keras_Tipps"
author: "Oliver and Beate"
date: "1/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tipps to use Keras 

```{r}
library(keras)
library(tensorflow)
library(tfprobability)
tf_version()
tf_probability()
# sessionInfo() # to see versions of all packages, also tf_probability
```
## Further resources
* Webpage https://keras.rstudio.com/index.html
* Notebooks to Manning book *Deep learning with R* https://github.com/jjallaire/deep-learning-with-r-notebooks
* R in colab: https://colab.research.google.com/notebook#create=true&language=r in the NB
  ```
    install.packages("keras")
    library(keras)
    library(tensorflow)
  ```

### Advanced
* Example calling python from within R: https://blogs.rstudio.com/tensorflow/posts/2019-12-10-variational-gaussian-process/

## Suprises

* We sometimes need to make sure that we have integers. E.g. `axis=1` gives a strange error message, because `1` is assigned to a numeric in R not an integer

* Axis are zero bases
```{r}
x = k_variable(matrix(rnorm(rep(0,3*4)),nrow = 3, ncol=4), dtype = 'float32')
#tf$reduce_mean(x,axis=1) #Does not work: NotFoundError: Could not find valid device for node
tf$reduce_mean(x,axis=0L) #Axis are zero based
str(1)
str(1L)
```

## Getting data to Keras (simple)

In the eager guide: https://keras.rstudio.com/articles/eager_guide.html they suggest to use tfdatasets. Probably better if you feed larger data (see later)
```{r}
(x = matrix(rnorm(rep(0,3*4)),nrow = 3, ncol=4))
(kc  = k_constant(x)) # turns x to tensor (not trainable)
(kt = k_variable(x))  # turns x to tf.variable (traninable)
(ka = keras_array(x, dtype='float32') )  # turns x to numpy array with can be used in TF
class(ka) #"numpy.ndarray"
```

## Getting data from Keras

In the eager mode
```{r}
(km=kt$numpy())  # turns the tf.variable kt to a matrix
```

## C




